---
title: "Gut Microbiota Report params$zoo"
author: "XXX, Uppsala University"
date: "January 7, 2025"
output: pdf_document
params:
  zoo: Helsinki
  div_df_path: 
---
```{r, echo=FALSE}
knitr::opts_chunk$set(echo = FALSE)
library(phyloseq)
library(ggplot2)

#plot_richness(ps, x = "Zoo", measures = c("Shannon", "Observed")) +
  #theme_bw()
```

```{r}
df_div <- data.frame(
  zoo = c("Nordens Ark", "Kolmården", "Skansen", 
          "Lycksele Zoo", "Orsa Predator Park",
          "Nordens Ark", "Kolmården", "Skansen",
          "Orsa Predator Park", "Orsa Predator Park"),
  individual = c("W1", "W2", "W3",
                 "W4", "W5",
                 "W6", "W7",
                 "W8", "W9", "W10"),
  alpha = round(runif(10, 2.5, 4.5), 3),
  beta  = round(runif(10, 0.1, 0.6), 3)
)
```

The present report summarizes the gut microbiota profiles of wolverines housed at your institution, based on high-throughput sequencing of the fecal samples you sent us. The results are intended to provide a baseline overview of microbial diversity, composition, and potential patterns observed across individuals. While microbiota data alone cannot diagnose health issues, it can serve as a complementary tool supporting husbandry decisions, dietary evaluations, and long-term welfare assessments.\\

We thank you for your collaboration and hope that the findings contribute to ongoing efforts to optimize care for captive wolverines and support collaborative knowledge-sharing across institutions. \\



#GRAF 1: Pie chart of composition: phylum and family


#GRAF 2: Functional analysis top 10

#Graf 3: NMDS plot of taxonomical distribution, the relevant individuals are circled. 

---
#his generates a sentence on the alpha and beta diversity based on a dataframe passed as params with the columns individual, zoo, alpha and beta
---

```{r}
write_microbiota_summary <- function(df, zoo_name) {
  
  # Filter to the zoo
  df_zoo <- df[df$zoo == zoo_name, ]
  
  indivs  <- df_zoo$individual
  alphas  <- df_zoo$alpha
  betas   <- df_zoo$beta
  
  n <- length(indivs)
  
  # --- Build alpha text ---
  if (n == 1) {
    alpha_text <- alphas
    
  } else if (n == 2) {
    alpha_text <- paste(alphas, collapse = " and ")
    
  } else {
    alpha_text <- paste(
      paste(alphas[-n], collapse = ", "),
      "and",
      alphas[n]
    )
  }
  
  # Ending for alpha: only add ", each." for >2 individuals
  ending_alpha <- ifelse(n > 2, ", each.", ".")
  
  alpha_sentence <- paste0(
    "Your individuals have an alpha diversity of ",
    alpha_text,
    ending_alpha
  )
  
  # --- Beta diversity summary ---
  # You can change this to unique(df_zoo$beta) if beta is per-zoo, not per-individual
  beta_value <- mean(df_zoo$beta, na.rm = TRUE)
  beta_value <- round(beta_value, 3)
  
  # Interpretation based on thresholds TA EV BORT DETTA 
  beta_comment <- dplyr::case_when(
    beta_value < 0.2 ~ "This indicates that the gut microbiota of your individuals is very similar.",
    beta_value < 0.4 ~ "This reflects a moderate level of difference between individuals.",
    TRUE             ~ "This suggests substantial variation in gut microbiota between your individuals."
  )
  
  beta_sentence <- paste0(
    "In total, your zoo has a beta diversity of ",
    beta_value,
    ". ",
    beta_comment
  )
  
  # --- Return both sentences as a list or full text block ---
  list(
    alpha_sentence = alpha_sentence,
    beta_sentence  = beta_sentence,
    full_text      = paste(alpha_sentence, beta_sentence)
  )
}
summary_text <- write_microbiota_summary(df_div, params$zoo)

```


`r summary_text$full_text`





