---
title: "Gut Microbiota Report `r params$zoo`"
author: "XXX, Uppsala University"
date: "January 7, 2025"
output: html_document
params:
  zoo: "Helsinki"
  
---
```{r, echo=FALSE, warning=FALSE, message=FALSE}
knitr::opts_chunk$set(echo = FALSE)
library(phyloseq)
library(ggplot2)
library(dplyr)
library(microbiome)
library(patchwork)


load("C:/Users/anton/Desktop/Applied Bioinformatics/Wolverine/scripts/report_generation/div_df.RData") 
div_df <- final_df 

load("C:/Users/anton/Desktop/Applied Bioinformatics/Wolverine/data/ps_filtered_1209.RData") 
ps <- ps_filtered 

load("C:/Users/anton/Desktop/Applied Bioinformatics/Wolverine/data/ps_rel_1209.RData")

load("C:/Users/anton/Desktop/Applied Bioinformatics/Wolverine/data/ps_merged_1201.RData")
ps_merged <- ps_merged_1201

df <- psmelt(ps_merged)

name_metadata <- read.csv(
  "C:/Users/anton/Desktop/Applied Bioinformatics/Wolverine/data/NGI_ID_to_Name.csv",
  sep = ";",
  stringsAsFactors = FALSE,
  header = TRUE
)

# Extract sample_data
sd <- sample_data(ps_merged)

# Map wolverine names
sd$Name <- name_metadata$Name[match(sd$NGI.ID, name_metadata$NGI.ID)]

# Update phyloseq object
sample_data(ps_merged) <- sd

# Select samples belonging to the target zoo
target_zoo <- as.character(params$zoo)

# Get sample names for the zoo
samples_in_zoo <- rownames(subset(sample_data(ps_merged), Zoo == target_zoo))
if(length(samples_in_zoo) == 0) stop(paste("No samples found for zoo:", target_zoo))

# Subset phyloseq object
ps_zoo <- prune_samples(samples_in_zoo, ps_merged)


#plot_richness(ps, x = "Zoo", measures = c("Shannon", "Observed")) +
  #theme_bw()
```


The present report summarizes the gut microbiota profiles of wolverines housed at your institution, based on high-throughput sequencing of the fecal samples you sent us. The results are intended to provide a baseline overview of microbial diversity, composition, and potential patterns observed across individuals. While microbiota data alone cannot diagnose health issues, it can serve as a complementary tool supporting husbandry decisions, dietary evaluations, and long-term welfare assessments.

We thank you for your collaboration and hope that the findings contribute to ongoing efforts to optimize care for captive wolverines and support collaborative knowledge-sharing across institutions. 

---
# Pie-Plots
---
```{r pie-function, echo=FALSE, warning=FALSE, message=FALSE}
make_pie_plot <- function(ps_object, top_n = 5) {

  # Transform to relative abundance
  ps_norm <- microbiome::transform(ps_object, "compositional")
  df <- psmelt(ps_norm)
  
  output_list <- list()
  sample_ids <- sample_names(ps_object)

  make_single_pie <- function(df_sub, tax_level, sample_label) {
    taxa_summary <- aggregate(Abundance ~ ., data=df_sub[c(tax_level, "Abundance")], sum)
    taxa_summary <- taxa_summary[order(-taxa_summary$Abundance), ]
    top_taxa <- as.character(taxa_summary[[tax_level]][1:min(nrow(taxa_summary), top_n)])

    df_sub[[tax_level]] <- as.character(df_sub[[tax_level]])
    df_sub[[tax_level]][!(df_sub[[tax_level]] %in% top_taxa)] <- "Other"
    df_sub[[tax_level]][is.na(df_sub[[tax_level]])] <- "Unassigned"

    df_clean <- aggregate(Abundance ~ ., data=df_sub[c(tax_level, "Abundance")], sum)
    df_clean$Percent <- df_clean$Abundance * 100

    legend_order <- c(top_taxa, "Unassigned", "Other")
    legend_order <- legend_order[legend_order %in% df_clean[[tax_level]]]
    df_clean[[tax_level]] <- factor(df_clean[[tax_level]], levels=legend_order)

    top_colors <- c("#009E73", "#56B4E9", "#F0E442", "#0072B2", "#D55E00", "#E69F00")
    tax_levels <- unique(df_clean[[tax_level]])
    tax_colors <- setNames(rep("#999999", length(tax_levels)), tax_levels)
    for(i in seq_along(top_taxa)) if(i <= length(top_colors)) tax_colors[top_taxa[i]] <- top_colors[i]
    if("Other" %in% tax_levels) tax_colors["Other"] <- "#999999"
    if("Unassigned" %in% tax_levels) tax_colors["Unassigned"] <- "#CC79A7"

    pie <- ggplot(df_clean, aes(x = "", y = Abundance, fill = !!sym(tax_level))) +
  geom_col(width = 1, color = "white") +
  coord_polar(theta = "y") +
  theme_void() +
  ggtitle(paste("Relative Abundance (", tax_level, "): ", sample_label, sep="")) +
  theme(plot.title = element_text(hjust = 0.5, size = 14),
        legend.title = element_blank()) +
  scale_fill_manual(values = tax_colors)


    return(pie)
  }

  # Generate pies per sample
  for(sid in sample_ids){
  df_sub <- df[df$Sample == sid, , drop = FALSE]
  
  # Get the wolverine name for this sample
  wolverine_name <- unique(df_sub$Name)  # should be one per sample
  
  pie_phylum <- make_single_pie(df_sub, "Phylum", wolverine_name)
  pie_family <- make_single_pie(df_sub, "Family", wolverine_name)
  combined <- pie_phylum + pie_family + patchwork::plot_layout(ncol = 2)

  output_list[[paste0(sid, "_Phylum")]] <- pie_phylum
  output_list[[paste0(sid, "_Family")]] <- pie_family
  output_list[[paste0(sid, "_Combined")]] <- combined
}



  return(output_list)
}

```
```{r pie-charts, fig.width=10, fig.height=6, fig.cap="Figure 1. Pie charts showing the composition of gut microbiota for each individual wolverine. For each individual, the left chart displays the relative abundance of bacteria at the phylum level, providing an overview of the main bacterial groups present. The right chart shows the family level, giving a more detailed view of which bacterial families make up the microbiota. The size of each slice represents the proportion of that group in the gut, allowing us to compare which bacteria are most common in each individual.", warning=FALSE, message=FALSE}
pies <- make_pie_plot(ps_zoo)

for(sid in names(pies)){
  if(grepl("_Combined$", sid)){
    print(pies[[sid]])
  }
}

```

---
#NMDS-plots:
---


```{r nmds-plot, echo=FALSE, warning=FALSE, message=FALSE, fig.width=8, fig.height=6, fig.cap="Figure 2. Ordination (NMDS) plot showing how similar or different the gut microbiomes are within each zoo. Each point represents an individual animal, and the distance between points reflects how different their microbiomes are: points that are close together indicate animals with similar gut bacteria, while those far apart have more distinct bacterial communities. The individuals from your are encircled in black.", fig.align='center'}
target_zoo <- as.character(params$zoo)

# Transform phyloseq object to relative abundance
ps.prop <- microbiome::transform(ps_merged, "compositional")

# Compute NMDS and fully suppress all console output
ord.nmds.bray <- suppressMessages(
  suppressWarnings(
    invisible(
      capture.output({
        ord <- ordinate(ps.prop, method = "NMDS", distance = "bray")
      }, type = "output")
    )
  )
)

# Use ord for plotting
p <- plot_ordination(ps.prop, ord, color = "Zoo", title = "Bray NMDS")

# Highlight target zoo samples
nmds_data <- plot_ordination(ps.prop, ord, color = "Zoo")$data
tmp <- nmds_data[nmds_data$Zoo == target_zoo, ]
tmp$SexLabel <- ifelse(tmp$Sex == 1, "Female",
                       ifelse(tmp$Sex == 0, "Male", "Unknown"))
tmp$Label <- tmp$Name  # use wolverine names

p +
  geom_point(
    data = tmp,
    aes(x = NMDS1, y = NMDS2),
    shape = 21,
    size = 10,
    stroke = 1.5,
    colour = "black",
    fill = NA,
    inherit.aes = FALSE
  ) +
  geom_text(
    data = tmp,
    aes(x = NMDS1, y = NMDS2, label = Label),
    size = 4,
    vjust = -2,
    hjust = 0.5,
    inherit.aes = FALSE
  )


```

---
#This generates a sentence on the alpha and beta diversity based on a dataframe passed as params with the columns individual, zoo, alpha and beta:
---

```{r}
write_microbiota_summary <- function(df, zoo_name) {
  
  df_zoo <- df[df$Zoo == zoo_name, ]
  
  indivs  <- df_zoo$Individual
  alphas  <- round(df_zoo$alpha,3)
  #betas   <- df_zoo$beta
  
  n <- length(indivs)
  
  if (n == 1) {
    alpha_text <- alphas
    
  } else if (n == 2) {
    alpha_text <- paste(alphas, collapse = " and ")
    
  } else {
    alpha_text <- paste(
      paste(alphas[-n], collapse = ", "),
      "and",
      alphas[n]
    )
  }
  
  ending_alpha <- ifelse(n > 2, ", each.", ".")
  
  alpha_sentence <- paste0(
    "Your individuals have an alpha diversity of ",
    alpha_text,
    ending_alpha
  )
  
  #beta_value <- mean(df_zoo$beta, na.rm = TRUE)
  #beta_value <- round(beta_value, 3)
  
  # Interpretation based on thresholds TA EV BORT DETTA 
  #beta_comment <- dplyr::case_when(
    #beta_value < 0.2 ~ "This indicates that the gut microbiota of your individuals is very similar.",
   # beta_value < 0.4 ~ "This reflects a moderate level of difference between individuals.",
   # TRUE             ~ "This suggests substantial variation in gut microbiota between your individuals."
#  )
  
 # beta_sentence <- paste0(
    #"In total, your zoo has a beta diversity of ",
   # beta_value,
   # ". ",
   # beta_comment
  #)
  
  list(
    alpha_sentence = alpha_sentence,
    #beta_sentence  = beta_sentence,
    full_text      = paste(alpha_sentence)
  )
}
summary_text <- write_microbiota_summary(div_df, params$zoo)

```


`r summary_text$full_text`